Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model

### 2023-06-19 src/docs_paper/paper.Rmd
Create figures.R and move from paper.Rmd

```{r echo = FALSE}
colours <- c("#56B4E9", "#009E73", "#E69F00")

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)
```

@@

```{r aghq, results='hide', fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension. Adaption occurs based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode. The integrand is $f(\\btheta) = \\text{sn}(0.3 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.5 \\theta_1 - 0.3 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
TMB::compile("2d.cpp")
dyn.load(TMB::dynlib("2d"))

obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")

grid <- expand.grid(
  theta1 = seq(-8, 8, length.out = 400),
  theta2 = seq(-8, 8, length.out = 400)
)

ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))

opt <- nlminb(
  start = obj$par,
  objective = obj$fn,
  gradient = obj$gr,
  control = list(iter.max = 1000, trace = 0)
)

sd_out <- TMB::sdreport(
  obj,
  par.fixed = opt$par,
  getJointPrecision = TRUE
)

mu <- opt$par
cov <- sd_out$cov.fixed

plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = "lightgrey") +
  coord_fixed(xlim = c(-8, 8), ylim = c(-8, 8), ratio = 1) +
  labs(x = "", y = "") +
  theme_minimal() +
  guides(size = FALSE) +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.y = element_blank()
  )

gg <- mvQuad::createNIGrid(2, "GHe", 3)

add_points <- function(plot0, gg) {

  points <- mvQuad::getNodes(gg) %>%
    as.data.frame() %>%
    mutate(weights = mvQuad::getWeights(gg))

  colnames(points) <- c("theta1", "theta2", "weights")

  plot0 +
    geom_point(
      data = points,
      aes(x = theta1, y = theta2, size = weights),
      alpha = 0.8,
      col = "#009E73",
      inherit.aes = FALSE
    ) +
    scale_size_continuous(range = c(1, 2))
}

plot1 <- add_points(plot0, gg) +
  labs(size = "")

#' Adapt by the mean
gg2 <- gg
mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 1)

plot2 <- add_points(plot0, gg2) +
  labs(size = "")

#' Adapt by the lower Cholesky
gg3 <- gg
mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)

plot3 <- add_points(plot0, gg3) +
  labs(size = "")

#' Adapt by the spectral
gg4 <- gg
mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)

plot4 <- add_points(plot0, gg4) +
  labs(size = "")

#' PCA-AGHQ
gg5 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
mvQuad::rescale(gg5, m = mu, C = cov, dec.type = 1)

plot5 <- add_points(plot0, gg5) +
  labs(size = "")

(plot1 + plot2) / (plot4 + plot5)
```

@@

rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% round(2)
rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% round(2)
rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% round(0)

rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% round(2)
rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% round(2)
rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% round(0)
```

@@

```{r ks, fig.height = 4, fig.cap="Example KS test for one parameter."}
df_compare <- readRDS("depends/beta_alpha.rds") %>%
  filter(method != "adam")

mean <- df_compare %>%
  filter(method == "tmbstan") %>%
  summarise(mean = mean(samples)) %>%
  pull(mean) %>%
  round(digits = 3)

sd <- df_compare %>%
  filter(method == "tmbstan") %>%
  summarise(sd = sd(samples)) %>%
  pull(sd) %>%
  round(digits = 3)

histogram <- df_compare %>%
  mutate(method = fct_recode(method, "PCA-AGHQ" = "aghq", "NUTS"= "tmbstan")) %>%
  ggplot(aes(x = samples, fill = method, col = method)) +
  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
  theme_minimal() +
  facet_grid(method~.) +
  labs(x = "beta_alpha", y = "Density", fill = "Method") +
  scale_color_manual(values = colours) +
  scale_fill_manual(values = colours) +
  theme(legend.position = "none")

grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), length.out = 1000)

tmb_ecdf <- stats::ecdf(filter(df_compare, method == "TMB") %>% pull(samples))
tmb_ecdf_df <- data.frame(x = grid, ecdf = tmb_ecdf(grid), method = "TMB")

aghq_ecdf <- stats::ecdf(filter(df_compare, method == "aghq") %>% pull(samples))
aghq_ecdf_df <- data.frame(x = grid, ecdf = aghq_ecdf(grid), method = "aghq")

tmbstan_ecdf <- stats::ecdf(filter(df_compare, method == "tmbstan") %>% pull(samples))
tmbstan_ecdf_df <- data.frame(x = grid, ecdf = tmbstan_ecdf(grid), method = "tmbstan")

# Add ECDF differences
tmb_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - tmb_ecdf_df$ecdf
aghq_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - aghq_ecdf_df$ecdf
tmbstan_ecdf_df$ecdf_diff <- 0

absmax <- function(x) x[which.max(abs(x))]

ks_tmb <- absmax(tmb_ecdf_df$ecdf_diff)
ks_aghq <- absmax(aghq_ecdf_df$ecdf_diff)

ecdf_df <- bind_rows(tmb_ecdf_df, aghq_ecdf_df, tmbstan_ecdf_df)

ecdf_df$method <- factor(ecdf_df$method, levels = c("TMB", "aghq", "tmbstan"))

ks_labeller <- function(x) toString(round(abs(x), 2))

ecdf_diff <- ggplot(ecdf_df, aes(x = x, y = ecdf_diff, col = method)) +
  geom_line() +
  geom_abline(intercept = ks_tmb, slope = 0, col = colours[1], linetype = "dashed", alpha = 0.8) +
  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = colours[1], alpha = 0.8) +
  geom_abline(intercept = ks_aghq, slope = 0, col = colours[2], linetype = "dashed", alpha = 0.8) +
  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = colours[2], alpha = 0.8) +
  scale_color_manual(values = colours) +
  labs(x = "beta_alpha", y = "ECDF difference") +
  guides(col = "none") +
  coord_cartesian(xlim = c(min(ecdf_df$x), max(ecdf_df$x)), clip = "off") +
  theme_minimal() +
  theme(plot.margin = unit(c(1, 3, 1, 1), "lines"))

histogram + ecdf_diff
```

@@

@@

Ambitious fast-track targets for scaling up ART treatment have been developed by UNAIDS, with the goal of "ending the AIDS epidemic by 2030".
Specifically, meeting the "90-90-90" fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
We found that for women both TMB and PCA-AGHQ underestimate these exceedance probabilities (Figure \ref{fig:exceedance}, first row).
We hypothesise that the sex related discrepancy occurs due to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.

@@

Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as greater than 1% incidence per year.
Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:exceedance}, second row).
This is surprising, in that we expect inferences from NUTS to be more heavy-tailed than those from TMB or PCA-AGHQ.

```{r exceedance, fig.cap="Both TMB and PCA-AGHQ are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities. For the second 90 target (top row) the inaccuracy varies substantially by sex."}
knitr::include_graphics("depends/exceedance.png")
```

>>>>

```{r echo = FALSE}
cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)

source("figures.R")
```

@@

```{r aghq, results='hide', fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension. Adaption occurs based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode. The integrand is $f(\\btheta) = \\text{sn}(0.3 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.5 \\theta_1 - 0.3 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
figA
```

@@

rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% signif(2)

rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% signif(0)
```

@@

```{r ks, fig.height = 4, fig.cap="Example KS test for one parameter."}
figB
```

@@

```{r}
exceedance <- read_csv("depends/exceedance.csv")

```

@@

Ambitious fast-track targets for scaling up ART treatment have been developed by UNAIDS, with the goal of "ending the AIDS epidemic by 2030".
Meeting the "90-90-90" fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
We found that for women both TMB and PCA-AGHQ underestimate these exceedance probabilities (Figure \ref{fig:exceedance}, first row).
We hypothesise this discrepancy in accuracy by sex is related to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.

@@

Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:exceedance}, second row).
This is surprising, in that we expect inferences from NUTS to be more heavy-tailed than those from TMB or PCA-AGHQ.

```{r exceedance, fig.cap="Though PCA-AGHQ does perform slightly better, both approximate inference methods are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities. For the second 90 target the inaccuracy varies substantially by sex."}
knitr::include_graphics("depends/exceedance.png")
```
